Functions for simulating data and designing studies of physiological flexibility in the acute glucocorticoid response to stressors

### 2021-09-03 3_r_scripts/Manuscript.Rmd
Changed the data simulation example from blackbirds to rats.

---
title: "Functions for simulating data and designing studies of physiological flexibility in the acute glucocorticoid response to stressors"
author: 
 - Conor C. Taff
header-includes:
  - \usepackage[font={footnotesize}, labelfont={bf}]{caption}
fontsize: 12pt
output: 
  bookdown::word_document2:
      number_sections: FALSE
      toc: FALSE
      extra_dependcies: ["flafter"]
  bookdown::pdf_document2:
      number_sections: FALSE
      toc: FALSE
      extra_dependencies: ["flafter"]
bibliography: references.bib
csl: animal-behaviour.csl
---

@@

Wild animals often experience unpredictable challenges that demand rapid and flexible responses. The glucocorticoid mediated stress response is one of the major systems that allows vertebrates to rapidly adjust their physiology and behavior. Given it's role in responding to challenges, evolutionary physiologists have focused on the consequences of between-individual and, more recently, within-individual variation in the acute glucocorticoid response. Although sophisticated approaches are available to partition this variation statistically, empirical studies of physiological flexibility are severely limited by the logistical challenges of measuring the same animal multiple times during a single acute response or across multiple instances of acute responses. Empiricists have largely adopted the strategy of standardizing sampling as much as possible to allow for comparison between individuals, but this standardization also makes it very difficult to detect certain types of variation in the functional shape of acute response curves. Data simulation is a powerful approach when empirical data are limited, but has not been adopted to date in studies of physiological flexibility. In this paper, I describe the `simcoRt` package, which includes functions that can generate realistic acute glucocorticoid response data with user specified characteristics. Simulated animals can be sampled continuously through an acute response and across as many separate responses as desired, while varying key parameters (e.g., the degree of correlation between the speed and scope of a response). Using this simulation, I explore several possible scenarios to highlight areas where simulation might either provide new insight into physiological flexibility directly or aid in designing empirical studies that are better able to test the hypotheses of interest. The package includes a vignette with details on use and code to reproduce all of the examples explored in this paper.

@@

The simulation is constructed as two main functions with several minor functions for downstream analysis. Detailed descriptions of the arguments to each function and a full vignette are included with the package documentation. Briefly, function `cort_sim1` samples the parameters shown in Figure \@ref(fig:concept-fig) from an arbitrary number of animals. These parameters are sampled from a multivariate normal distribution with user specified mean, variance, and covariance for each parameter. I consider these values to be the 'true', unobserved, phenotype of the animal (setting aside the question of whether or not a 'true' physiological phenotype exists).

@@

The goal of this simulation is to provide a flexible tool that can produce realistic datasets of physiological flexibility for a variety of different systems and scenarios. As such, there are many possible applications and here I briefly highlight a few possibilities. These are by no means exhaustive, and I hope the simulation will be a useful tool to guide empirical work for specific hypotheses and study systems. Within each scenario, I have illustrated how the simulation functions might be used to address the particular question of interest, but I have not fully explored all the possible permutations of parameters systematically, because these will depend to a large extent on the empirical details of the system being studied.

@@

Here, I compare simulated data to an empirical dataset from red-winged blackbirds (*Agelaius phoeniceus*) to demonstrate how `simcoRt` can create synthetic datasets similar to observed data. This dataset is included in the `simcoRt` package and is fully described in the documentation. Briefly, acute glucocorticoid responses were measured using a 7-point sampling scheme, with corticosterone measured at 1, 3, 7, 15, 30, 45, and 60 minutes post capture. I simulated data using the functions described above starting with the input values calculated directly from the empirical data. The simulation creates a new dataset that has similar variation and patterns to empirical data (Figure \@ref(fig:rwbb-demo)A) along with a population wide corticosterone response curve shape that closely matches the empirical data (Figure \@ref(fig:rwbb-demo)B). In this case, the plotted simulation data include the same number of animals sampled at the same time points as the empirical data, but these sampling points and total sample size can easily be changed as desired. The parameterized simulation can now be used to test the sensitivity of any number of experimental designs before additional data is collected.

```{r rwbb-demo, echo = FALSE, fig.height = 3.5, fig.width = 8, message = FALSE, warning = FALSE, fig.cap = "Panel A shows the acute corticosterone response for measured (blue) or simulated (orange) red-winged blackbirds measured at seven time points. Panel B shows the mean and standard error of the two datasets."}
      rwde <- subset(rwbb, rwbb$lh_stg == "early-breeding")  
      rwde <- subset(rwde, rwde$id != "452")   # This one has an unusually high cort response
    set.seed(165)
        sim_rwd <- cort_sim2(cort_sim1(
       n = 23,
       base_mu = 3.5,
       base_sd = 3.5,
       slope_mu = 17,
       slope_sd = 0.7,
       fastpct_mu = 0.7,
       fastpct_sd = 0.03,
       max_mu = log(56),
       max_sd = 0.5,
       speed_mu = 65,
       speed_sd = 5,
       return_mu = 120,
       maxtime_mu = 25,
       cor_max_slope = -0.7
     ), 
     sample_times = 1,
     bleed_times = c(1, 3, 8, 15, 30, 45, 60),
     assay_error = 0.2,
     sm_span = 0.2)

     sim_rwd$simulated_dataset_long$type <- "simulated"
     sim_rwd2 <- sim_rwd$simulated_dataset_long[, c("animal_sample", "time", "cort", "type")]
     rwde2 <- rwde[, c("id", "lat_sec", "cort")]
     rwde2$lat_sec <- rwde2$lat_sec / 60
     rwde2$type <- "red-winged blackbird"
     colnames(rwde2) <- c("animal_sample", "time", "cort", "type")

     comp_sim <- rbind(rwde2, sim_rwd2)

     p1 <- ggplot(comp_sim, mapping = aes(x = time, y = cort, by = animal_sample, color = type)) +
       geom_line(alpha = 0.7, size = 0.7) +
       theme_classic() +
       coord_cartesian(xlim = c(0, 60), ylim = c(0, 125)) +
       xlab("Time (minutes)") + ylab("Corticosterone (ng/\U00B5l)") +
       scale_color_manual(values = c("slateblue", "orange")) +
       guides(color = FALSE) +
       theme(axis.title = element_text(size = 14)) +
       annotate(geom = "text", x = -Inf, y = Inf, hjust = -.5, vjust = 1.5, label = "A")

     p2 <- ggplot(comp_sim, mapping = aes(x = time, y = cort, color = type, fill = type)) +
       geom_smooth() +
       theme_classic() + 
       coord_cartesian(xlim = c(0, 60), ylim = c(0, 125)) +
       xlab("Time (minutes)") + ylab("Corticosterone (ng/\U00B5l)") +
       scale_color_manual(values = c("slateblue", "orange")) +
       scale_fill_manual(values = c("slateblue", "orange")) +
       theme(legend.position = c(0.7, 0.15), axis.title = element_text(size = 14)) +
       guides(color = guide_legend(title = ""), fill = guide_legend(title = "")) +
       annotate(geom = "text", x = -Inf, y = Inf, hjust = -.5, vjust = 1.5, label = "B")

@@

@@

Several patterns can be identified by examining the results of this simulation. First, the correlation between the true maximum glucorticoid value and fitness does not differ for populations simulated with high or low between-individual variation (figure \@ref(fig:fitness-ex) A-D). In all cases, however, the observed correlation is always lower than the true correlation and always lowest in the population with low between-individual variation. The ubquity of this pattern is a product of the simulation structure, because adding measurement error or within-individual variation effecively add noise to the true correlation. It is important to note that in the real world, it is unlikely that this pattern would be so universal, because unmeasured variables could influence both fitness and glucocorticoids. For example, if habitat quality directly alters fitness and glucocorticoids, the observed correlation could be stronger than the 'true' correlation. Thus, interpretation of these results should be made cautiosly in light of the simplicity of the simulation compared to real world conditions.

@@

The fact that low between-individual variation in maximum glucocorticoids makes it harder to detect true glucocorticoid-fitness relationships across a wide range of conditions has important consequences for interpreting empirical results.

@@

One of the major benefits of simulating glucocorticoid response curves will be the ability to design optimal sampling strategies before data are collected. A simulation can be constrained to match any real world limitations (e.g., maximum number of samples possible per individual) and then explored to determine how to best allocate sampling resources. The specifics of this task will vary considerably with the study system and question being addressed, but here I illustrate one possible application. Consider an experiment in which the acute glucocorticoid response of a treatment group and control group are compared after some experimental manipulation. The details of the manipulation are unimportant here, but suppose that the prediction is that this manipulation should result in a difference in the speed of the corticosterone response between our two groups, such that the treatment group will reach it's maximum glucocorticoid value faster than the control group (figure \@ref(fig:prediction-ex)).

@@

#plot(1,1)

@@

**Still working on thiis section. Will have two treatment groups with different response shapes and evaluate how well different sampling schemes recover those shapes.**

### Repeatability?

The other example I had thought about including was something showing the repeatability functions and how repeatability of different measures might differ, but I'm leaning towards not including this since there is already a lot here.

@@

>>>>

---
title: "Functions for simulating data and designing studies of physiological flexibility in the acute glucocorticoid response to stressors"
author: 
 - Conor C. Taff
header-includes:
  - \usepackage[font={footnotesize}, labelfont={bf}]{caption}
fontsize: 12pt
output: 
  bookdown::pdf_document2:
      number_sections: FALSE
      toc: FALSE
      extra_dependencies: ["flafter"]
  bookdown::word_document2:
      number_sections: FALSE
      toc: FALSE
      extra_dependcies: ["flafter"]
bibliography: references.bib
csl: animal-behaviour.csl
---

@@

Wild animals often experience unpredictable challenges that demand rapid and flexible responses. The glucocorticoid mediated stress response is one of the major systems that allows vertebrates to rapidly adjust their physiology and behavior. Given it's role in responding to challenges, evolutionary physiologists have focused on the consequences of between-individual and, more recently, within-individual variation in the acute glucocorticoid response. Although sophisticated approaches are available to partition this variation statistically, empirical studies of physiological flexibility are severely limited by the logistical challenges of measuring the same animal multiple times during a single acute response or across multiple instances of acute responses. Empiricists have largely adopted the strategy of standardizing sampling as much as possible to allow for comparison between individuals, but this standardization also makes it very difficult to detect certain types of variation in the functional shape of acute response curves. Data simulation is a powerful approach when empirical data are limited, but has not been adopted to date in studies of physiological flexibility. In this paper, I describe the `simcoRt` package, which includes functions that can generate realistic acute glucocorticoid response data with user specified characteristics. Simulated animals can be sampled continuously through an acute response and across as many separate responses as desired, while varying key parameters (e.g., the degree of correlation between the speed and scope of a response). Using this simulation, I explore several possible scenarios to highlight areas where simulation might either provide new insight into physiological flexibility directly or aid in designing empirical studies that are better able to test the hypotheses of interest. All of the code required to reproduce the examples presented here is included in a public GitHub repository.

@@

The simulation is constructed as two main functions with several minor functions for downstream analysis. Detailed descriptions of the arguments to each function are included with the package documentation. Briefly, function `cort_sim1` samples the parameters shown in Figure \@ref(fig:concept-fig) from an arbitrary number of animals. These parameters are sampled from a multivariate normal distribution with user specified mean, variance, and covariance for each parameter. I consider these values to be the 'true', unobserved, phenotype of the animal (setting aside the question of whether or not a 'true' physiological phenotype exists).

@@

The goal of this simulation is to provide a flexible tool that can produce realistic datasets of physiological flexibility for a variety of different systems and scenarios. As such, there are many possible applications and here I briefly highlight a few possibilities. These are by no means exhaustive, and I hope the simulation will be a useful tool to guide empirical work for specific hypotheses and study systems. Within each scenario, I have illustrated how the simulation functions might be used to address the particular question of interest, but I have not fully explored all the possible permutations of parameters systematically, because these will depend to a large extent on the empirical details of the system being studied. A complete set of reproducible code to create all of the examples presented in this paper is available on GitHub (https://github.com/cct663/speed_vs_scope).

@@

```{r kool-demo, echo = FALSE, fig.height = 3.5, fig.width = 8, message = FALSE, warning = FALSE, fig.cap = "Panel A shows the acute corticosterone response for measured (blue) or simulated (orange) rats measured at five time points. Panel B shows the mean and standard error of the two datasets. Empirical data are extracted from Koolhaas et al. 2010 Figure 6 using the WebPlotDigitizer tool."}
     # Load and set up data from Koolhaas et al 2010
#   kool <- read.delim(here::here("1_raw_data/koolhaas_2010_figure6_data.txt"))
#   kool <- subset(kool, kool$time > -2)
#   kool$type <- "Koolhaas et al. 2010"
# 
# # Create simulated dataset with similar characteristics  
#     set.seed(175)
#           sim_kool <- cort_sim2(cort_sim1(
#            n = 14,
#            base_mu = 11.6,
#            base_sd = 3.65,
#            slope_mu = 3.5,
#            slope_sd = 0.9,
#            fastpct_mu = 0.9,
#            fastpct_sd = 0.1,
#            max_mu = log(24),
#            max_sd = 0.1,
#            speed_mu = 12,
#            speed_sd = 2.5,
#            return_mu = 42,
#            maxtime_mu = 1,
#            maxtime_sd = 0.5
#          ), 
#          sample_times = 1,
#          bleed_times = c(0, 1, 5, 15, 30, 60),
#          assay_error = 0.1,
#          sm_span = 0.2)  
#       
# # Combine simulated and real data
#     sim_kool$simulated_dataset_long$type <- "simulated"
#      sim_kool2 <- sim_kool$simulated_dataset_long[, c("animal_sample", "time", "cort", "type")]
#      kool2 <- sim_kool2[, c("animal_sample", "time", "cort")]
#      kool2$type <- "simulated"
#      colnames(kool2) <- c("animal_sample", "time", "cort", "type")
#      
#      kool <- kool[, c("animal", "time", "cort", "type")]
#      colnames(kool)[1] <- "animal_sample"
#      comp_kool <- rbind(kool, kool2)

# save the data to file and load back in
     #saveRDS(comp_kool, here::here("5_other_outputs/comp_kool.rds"))
     comp_kool <- readRDS(here::here("5_other_outputs/comp_kool.rds"))

# Plot the data
    p1 <- ggplot(comp_kool, mapping = aes(x = time, y = cort, by = animal_sample, color = type)) +
       geom_line(alpha = 0.7, size = 0.7) +
       theme_classic() +
       coord_cartesian(xlim = c(0, 60), ylim = c(0, 45)) +
       xlab("Time (minutes)") + ylab("Corticosterone (\U00B5g/dl)") +
       scale_color_manual(values = c("slateblue", "orange")) +
       guides(color = "none") +
       theme(axis.title = element_text(size = 14)) +
       annotate(geom = "text", x = -Inf, y = Inf, hjust = -.5, vjust = 1.5, label = "A")

     p2 <- ggplot(comp_kool, mapping = aes(x = time, y = cort, color = type, fill = type)) +
       geom_smooth() +
       theme_classic() + 
       coord_cartesian(xlim = c(0, 60), ylim = c(0, 45)) +
       xlab("Time (minutes)") + ylab("Corticosterone (\U00B5g/dl)") +
       scale_color_manual(values = c("slateblue", "orange")) +
       scale_fill_manual(values = c("slateblue", "orange")) +
       theme(legend.position = c(0.7, 0.15), axis.title = element_text(size = 14)) +
       guides(color = guide_legend(title = ""), fill = guide_legend(title = "")) +
       annotate(geom = "text", x = -Inf, y = Inf, hjust = -.5, vjust = 1.5, label = "B")

@@

To demonstrate this utility, I have redrawn data from Koolhaas et al. [-@koolhaas2010]. As part of that study, a series of corticosterone measurements were collected during and after an acute stressor from 14 laboratory rats *Rattus norvegicus* using permanently implanted jugular vein canulae. I next simulated data using the functions described above starting with the input values calculated directly from the empirical data. The simulation creates a new dataset that has similar variation and patterns to the empirical data (Figure \@ref(fig:kool-demo)A) along with a population wide corticosterone response curve shape that closely matches the empirical data (Figure \@ref(fig:kool-demo)B). In this case, the plotted simulation data include the same number of animals sampled at the same time points as the empirical data, but these sampling points and total sample size can easily be changed as desired. The parameterized simulation can now be used to test the sensitivity of any number of experimental designs before additional data is collected.

@@

Several patterns can be identified by examining the results of this simulation. First, the correlation between the true maximum glucocorticoid value and fitness does not differ for populations simulated with high or low between-individual variation (figure \@ref(fig:fitness-ex) A-D). In all cases, however, the observed correlation is always lower than the true correlation and always lowest in the population with low between-individual variation. The ubiquity of this pattern is a product of the simulation structure, because adding measurement error or within-individual variation effectively adds noise to the true correlation. It is important to note that in the real world, it is unlikely that this pattern would be so universal, because unmeasured variables could influence both fitness and glucocorticoids. For example, if habitat quality directly alters fitness and glucocorticoids, the observed correlation could be stronger than the 'true' correlation. Thus, interpretation of these results should be made cautiosly in light of the simplicity of the simulation compared to real world conditions.

@@

The fact that low between-individual variation in maximum glucocorticoids makes it harder to detect true glucocorticoid-fitness relationships across a wide range of conditions has important consequences for interpreting empirical results. Many studies have demonstrated different relationships (or lack thereof) between corticosterone and fitness at different life history stages [add our FEE & Bonier papers CITE], but it is also well known that the absolute amount of between individual variation in glucocorticoid traits varies considerably at different stages [CITE]. Our simulation demonstrates that the power to detect true relationships will differ drastically across these conditions even with similar study designs and samples sizes, suggesting that great care is needed to conclusively demonstrate different glucocorticoid-fitness relationships across contexts.

@@

One of the major benefits of simulating glucocorticoid response curves will be the ability to design optimal sampling strategies before data are collected. A simulation can be constrained to match any real world limitations (e.g., maximum number of samples possible per individual) and then explored to determine how to best allocate sampling resources. The specifics of this task will vary considerably with the study system and question being addressed, but here I illustrate one possible application. Consider an experiment in which the acute glucocorticoid response of a treatment group and control group are compared after some experimental manipulation. The details of the manipulation are unimportant here, but suppose that the prediction is that this manipulation should result in a difference in the speed of the corticosterone response between our two groups, such that the treatment group will reach it's maximum glucocorticoid value faster than the control group, but will not differ in the maximum value itself (figure \@ref(fig:prediction-ex)).

@@

set.seed(295)

# Simulate group A
        sim_grpa <- cort_sim2(cort_sim1(
       n = 20,
       slope_mu = 9,
       fastpct_mu = 0.7,
       max_sd = 0.14,
       speed_mu = 20,
       speed_sd = 4
     ), 
     sample_times = 1,
     sm_span = 0.2,
     bleed_times = c(2, 30))

# Simulate group B
        sim_grpb <- cort_sim2(cort_sim1(
       n = 20,
       slope_mu = 11,
       fastpct_mu = 0.55,
       max_sd = 0.14,
       speed_mu = 32,
       speed_sd = 4
     ), 
     sample_times = 1,
     sm_span = 0.2,
     bleed_times = c(2, 30))

# Combine the two datasets
      sim_grpa$timecourse_long$type <- "Treatment"
      sim_a2 <- sim_grpa$timecourse_long[, c("animal_sample", "time", "cort", "type")]
      sim_grpb$timecourse_long$type <- "Control"
      sim_b2 <- sim_grpb$timecourse_long[, c("animal_sample", "time", "cort", "type")]

      comp_sim2 <- rbind(sim_a2, sim_b2)

     p1 <- ggplot(comp_sim2, mapping = aes(x = time, y = cort, by = animal_sample, color = type)) +
       geom_line(alpha = 0.3, size = 0.7) +
       theme_classic() +
       coord_cartesian(xlim = c(0, 60), ylim = c(0, 75)) +
       xlab("Time (minutes)") + ylab("Corticosterone (ng/\U00B5l)") +
       scale_color_manual(values = c("slateblue", "orange")) +
       #guides(color = "none") +
       theme(axis.title = element_text(size = 14)) +
       annotate(geom = "text", x = -Inf, y = Inf, hjust = -.5, vjust = 1.5, label = "A") +
       geom_smooth(se = FALSE, size = 1.5, mapping = aes(by = type, color = type)) +
       theme(legend.title = element_blank())

      sim_grpa$true_values$type <- "Treatment"
      sim_a22 <- sim_grpa$true_values[, c("baseline", "slope", "fastpct", "speed", "max", "atmax", "type")]
      sim_grpb$true_values$type <- "Control"
      sim_b22 <- sim_grpb$true_values[, c("baseline", "slope", "fastpct", "speed", "max", "atmax", "type")]

      comp_sim3 <- rbind(sim_a22, sim_b22)
      comp_sim4 <- pivot_longer(comp_sim3, cols = c("baseline", "slope", "fastpct", "speed", "max", "atmax"), names_to = "trait", values_to = "value")

      ggplot(comp_sim4, mapping = aes(x = type, y = value, color = type, fill = type)) +
        geom_boxplot(alpha = 0.5) +
        scale_color_manual(values = c("slateblue", "orange")) +
       scale_fill_manual(values = c("slateblue", "orange")) +
        facet_wrap(~ trait, nrow = 1, scales = "free") +
        xlab("") +
        theme_bw() +
        theme(axis.text.x = element_blank())

     p2 <- ggplot(comp_sim2, mapping = aes(x = time, y = cort, color = type, fill = type)) +
       geom_smooth() +
       theme_classic() + 
       coord_cartesian(xlim = c(0, 60), ylim = c(0, 125)) +
       xlab("Time (minutes)") + ylab("Corticosterone (ng/\U00B5l)") +
       scale_color_manual(values = c("slateblue", "orange")) +
       scale_fill_manual(values = c("slateblue", "orange")) +
       theme(legend.position = c(0.7, 0.15), axis.title = element_text(size = 14)) +
       guides(color = guide_legend(title = ""), fill = guide_legend(title = "")) +
       annotate(geom = "text", x = -Inf, y = Inf, hjust = -.5, vjust = 1.5, label = "B")

      ggpubr::ggarrange(p1, p2, nrow = 1)

@@

@@

How to interpret some of the empirical studies that are published in light of simulations?
